An Analysis of Financial Contributions to Presidential Campaigns in Illinois with R - By Tiffany Li
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(stringr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("~/Desktop/Udacity Nanodegree/P3/data set")
getwd()
PC <- read.csv('P1-IL-2.csv', header = TRUE, fill = TRUE)
```

#### Now that we have successfully loaded the dataset and packages, we can take a lot at relevant information about data, such as the structure, data types, column names. We can also display the first few rows to get a better sense of the data.
#### Also, we can find the description of variables in this linke: http://www.fec.gov/finance/disclosure/metadata/DataDictionaryCommitteeMaster.shtml.
```{r}
# Exam the dataset
str(PC)
names(PC)
head(PC)
```

#### By looking at the data types in the str() function, I detected 5 columns whose data types need to be converted. 'Index' is just an id so we should change it from integer to character. 'Contbr_zip' represents the zipcode of contributor so it should be factor instead of integer. 'Contb_receipt_amt' is the amount contributed therefore needs to be represented in numeric form instead of factor. 'File_num' is the file number and should be converted from integer to factor. Lastly, I think character is a better way than factor to represent 'Tran_id', which is transaction ID and has unique value for each row.
```{r}
# Convert data types if neccessary
PC$index <- as.character(PC$index)
PC$contbr_zip <- as.factor(PC$contbr_zip)
PC$contb_receipt_amt <- as.numeric(as.character(PC$contb_receipt_amt))
PC$contb_receipt_dt <- as.Date(PC$contb_receipt_dt, format = "%d-%b-%y")
PC$file_num <- as.factor(PC$file_num)
PC$tran_id <- as.character(PC$tran_id)
```


#### Now that we have all the correct data types, we can look at the summary statistics to get a rough idea of the data.
```{r}
# Exam average contribution for each candidate
summary(PC$cand_nm)
summary(PC$contbr_nm)
summary(PC$contbr_city)
summary(PC$contbr_zip)
summary(PC$contbr_employer)
summary(PC$contbr_occupation)
summary(PC$contb_receipt_amt)
summary(PC$contb_receipt_dt)
summary(PC$contb_receipt_desc)
summary(PC$memo_cd)
summary(PC$memo_text)
summary(PC$election_tp)
```
#### By simply looking at the summary statistics, we already learnt a great amount of information about our dataset. We could see how many contributions each candidate received, how many contributions were from each contributor, city, zip code, emoployer and occupation, summary statistics of contribution amount, the dates of contribution, amount of contributions made to primary and general election. Lastly, from looking at the memo we can distinguish the earmarked contribution, super-pac, in-kind donation, reattribution/redesignation and refunds.

#### Previously we saw that there are some candidates with fewer than 10 contributions. As a result of very few contributions, plots or summary statistics for these candidate might be misleading because of the small data size. Therefore I will exclude these candidate for the rest of the analysis.
```{r}
# Exclude candidates with fewer than 10 contributions
summary(PC$cand_nm)

PC <- subset(PC, !(PC$cand_nm == "O'Malley, Martin Joseph" | PC$cand_nm == "Graham, Lindsey O." | PC$cand_nm == "Jindal, Bobby" | PC$cand_nm == "Santorum, Richard J." | PC$cand_nm == "McMullin, Evan")) 
                    
```

### Create new variable
```{r}
# Create a new variable for candidate's associated political party
PC$cand_nm_party <- 'Republican'
```

# Univariate Plots Section
### Let's understand the data by visualizations of single variable.
#### How many contributions (by transaction) did each candidate receive?
```{r echo=FALSE, Univariate_Plots}
# Bar plot of total contributions (by transaction) for each candidate
theme_set(theme_gray(base_size = 10))
reorder_size <- function(x) {
  factor(x, levels = names(sort(table(x), decreasing = FALSE)))
}

ggplot(aes(x = reorder_size(cand_nm)), data = PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```
#### In terms of contribution transactions, Clinton received the most contributions, followed by Sanders, Cruz and Trump. Notice that we added a log10 scale to make the plot easier to read because the results vary significantly across candidates.

#### How many contribution transactions are from each city in Illinois?
```{r echo=FALSE, Univariate_Plots}
# Bar plot of total contributions (by transaction) from each city
ggplot(aes(x = contbr_city), data = PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```
#### We can observe that Chicago has the most contributions. However, the large amount of city names makes the plot hard to read.

#### How many contribution transactions are from each contributor employer?
```{r echo=FALSE, Univariate_Plots}
ggplot(aes(x = contbr_employer), data = PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```
#### We obtained a plot that represents the number of contributions from each employer. But, similarly, having many employers in the dataset makes the plot hard to read.


# Univariate Analysis
### What is the structure of your dataset?
```{r}
# Display the structure of the dataset
str(PC)
```


### What is/are the main feature(s) of interest in your dataset?
#### I am mainly interested in the following features: candidate name, contribution receipt amount, since the most interesting topic is analyzing the amount of contributions received by candidates.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
#### Features like contributor name, contributor city, contributor zipcode, contributor employer, contributor occupation, contribution receipt date, election type can provide further insights of the dataset. We can find relationship between these features and the features I mentioned in the last question. For example, we can look at the amount of contributions from different cities, or tries to determine if occupation has an effect on contribution amount.

### Did you create any new variables from existing variables in the dataset?
#### I created a new variable cand_nm_party, which is the candidate's name plus their party information(e.g: Democrat, Republican etc) in order to understand the differences among different party.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
#### .


# Bivariate Plots Section
### We can investigate the dataset by looking at two variables. 
#### First, let's find out the total amount of contributions received by each candidate.
```{r echo=FALSE, Bivariate_Plots}
# Visualize total contribution amount for each candidate
ggplot(aes(x = cand_nm, y = contb_receipt_amt), fill = election_tp, data = PC) +
  stat_summary_bin(aes(y = reorder(contb_receipt_amt)), fun.y = 'sum', geom = 'bar') +
  coord_flip()

ggplot(data=PC, aes(x=reorder(cand_nm, contb_receipt_amt, function(x){ sum(x) }), 
                    y=contb_receipt_amt)) +
geom_bar(stat="summary", fun.y=sum) +
coord_flip()

?reorder
```
#### Similar to the univariate plot of contribution transactions of each candidate, we are seeing that Clinton received the highest total amount of contribution, followed by Sanders, Cruz and Trump. 

#### We can also exam the median amount of contributions received by each candidate, which might indicate chracteristics about their supporters.
```{r echo=FALSE, Bivariate_Plots}
# Visualize the median contribution amount for each candidate
ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = PC) +
  stat_summary_bin(aes(y = contb_receipt_amt), fun.y = 'median', geom = 'bar') +
  coord_flip()
```
#### It appears that the candidates with high median contribution amount also had fewer contributions. Therefore this information might be misleading due to the small data size. It would be helpful to add number of contributions received to this plot.

#### Boxplots of contribution amount are also helpful, as they show show 25%, 50% and 75% percentile of the contribution amount for each candidate, along with outliers.
```{r echo=FALSE, Bivariate_Plots}
# Visualize boxplot for amount of contributions received for each candidate
ggplot(aes(x = cand_nm, y = contb_receipt_amt, fill = election_tp), data = PC) +
  geom_boxplot() +
  coord_flip() +
  ylim(0, 6000) +
  scale_y_log10()
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection