An Analysis of Financial Contributions to Presidential Campaigns in Illinois with R - By Tiffany Li
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

# install.packages("knitr", dependencies = T, repos="http://cran.rstudio.com/")
library(knitr)
library(ggplot2)
require(ggplot2)
library(dplyr)
library(stringr)
library(knitr)
```

```{r global_options, include=FALSE}
# Set the global options so that warnings and notifications don't appear in knit HTML file
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("~/Desktop/Udacity Nanodegree/P4/data set")
getwd()
PC <- read.csv('P1-IL-2.csv', header = TRUE, fill = TRUE)
```

#### Now that we have successfully loaded the dataset and packages, we can take a look at a lot of relevant information about data, such as the structure, data types, column names. We can also display the first few rows to get a better sense of the data.
#### Also, the description of variables can be found in this link: ftp://ftp.fec.gov/FEC/Presidential_Map/2016/DATA_DICTIONARIES/CONTRIBUTOR_FORMAT.txt
```{r}
# Exam the dataset
str(PC)
names(PC)
head(PC)
```

#### By looking at the data types in the str() function, I detected 6 columns whose data types need to be converted. 'index' is just an id so we should change it from integer to character. 'contbr_zip' represents the zipcode of contributor so it should be factor instead of integer. 'contb_receipt_amt' is the amount contributed therefore needs to be represented in numeric form instead of factor. 'contb_receipt_dt' is the date when contribution was received and should be represented in date form. 'file_num' is the file number and should be converted from integer to factor. Lastly, I think character is a better way than factor to represent 'tran_id', which is transaction ID and has unique value for each row.
```{r}
# Convert data types if neccessary
PC$index <- as.character(PC$index)
PC$contbr_zip <- as.factor(PC$contbr_zip)
PC$contb_receipt_amt <- as.numeric(as.character(PC$contb_receipt_amt))
PC$contb_receipt_dt <- as.Date(PC$contb_receipt_dt, format = "%d-%b-%y")
PC$file_num <- as.factor(PC$file_num)
PC$tran_id <- as.character(PC$tran_id)
```


#### Now that we have all the correct data types, we can begin the analysis.

#### There are some candidates with very few contributions compared to others. As a result, plots or summary statistics for these candidate might be misleading because of the small data size. Therefore I will exclude these candidate for the rest of the analysis.
```{r}
# Exclude candidates with fewer than 10 contributions
PC <- subset(PC, !(PC$cand_nm == "O'Malley, Martin Joseph" |
                     PC$cand_nm == "Graham, Lindsey O." |
                     PC$cand_nm == "Jindal, Bobby" |
                     PC$cand_nm == "Santorum, Richard J." |
                     PC$cand_nm == "McMullin, Evan")) 
```

#### It is also perhaps helpful to add the information about the political parties associated with the candidates.
```{r}
# Create a new variable for candidates' associated political parties by creating a new function rnParty and applying that function

Democratic <- c('Clinton, Hillary Rodham', 
                'Lessig, Lawrence', 
                'Sanders, Bernard')

rnParty <- function(x) {
if(is.na(x)){
  NA
}else if(x %in% Democratic){
  'D'
}else if(x == 'Johnson, Gary'){
  'L'
}else if(x == 'Stein, Jill'){
  'G'
}else{
  'R'
}
}
   
PC$cand_nm_party <- apply(PC['cand_nm'],1,rnParty)
```

# Univariate Plots Section
### Let's understand the data by visualizations of single variable.
#### How many contributions (by transaction) did each candidate receive?
```{r echo=FALSE, Univariate_Plots_1}
# Bar plot of total contributions (by transaction) for each candidate in descending order
reorder_size <- function(x) {
factor(x, levels = names(sort(table(x), decreasing = FALSE)))
}

ggplot(aes(x = reorder_size(cand_nm)), data = PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```

#### In terms of contribution transactions, Clinton received the most contributions, followed by Sanders, Cruz and Trump. Notice that we added a log10 scale to make the plot easier to read because the results vary significantly across candidates.

#### How many contribution transactions are from each city in Illinois?
```{r echo=FALSE, Univariate_Plots_2}
# Bar plot of total contributions (by transaction) from each city in descending order
ggplot(aes(x = reorder_size(contbr_city)), data =PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```

#### We can observe that Chicago has the most contributions. However, the large amount of city names makes the plot hard to read.

#### How many contribution transactions are from each contributor employer?
```{r echo=FALSE, Univariate_Plots_3}
ggplot(aes(x = reorder_size(contbr_employer)), data = PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```

#### We obtained a plot that represents the number of contributions from each employer. But, similarly, having many employers in the dataset makes the plot hard to read.

#### How many contribution transactions are from each occupation?
```{r echo=FALSE, Univariate_Plots_4}
ggplot(aes(x = reorder_size(contbr_occupation)), data = PC) +
  geom_bar() +
  coord_flip() +
  scale_y_log10()
```

#### We obtained a plot that represents the number of contributions from each contributor's occupation. But, similarly, having many occupations in the dataset makes the plot hard to read.

#### What is first quantile, median and third quantile of the contribution amount?
```{r echo=FALSE, Univariate_Plots_5}
# Visualize the contribution amount for all candidates using boxplot
summary(PC$contb_receipt_amt)

ggplot(aes(x = 1, y = contb_receipt_amt), data = PC) +
  geom_boxplot() +
  scale_y_log10()
```

#### The first quantile of contribution is $15, median is $27 and the third quantile is $100.

#### How many contributions were made to primary election and general election?
```{r echo=FALSE, Univariate_Plots_6}
# Display the number of contributions made to general and primary elections

ggplot(aes(x = election_tp), data = PC) +
  geom_bar()
```

#### We can see that 44,923 contributions were made to primary elections and 20,545 contributions were made to general elections.

#### It is apparent that more contributions were made to primary elections than general elections.

#### How many contributions were made to different political parties?
```{r echo=FALSE, Univariate_Plots_7}
# Visualize the number of contributions made to different political parties using bar plots
ggplot(aes(x = cand_nm_party), data = PC) +
geom_bar()
```

#### We can see that the Democratic party, despite having fewer candidates, received the most contributions, followed by the Republican party. Green party and libertarian party received much fewer contributions by comparison.

# Univariate Analysis
### What is the structure of your dataset?
```{r}
# Display the structure of the dataset
str(PC)
```


### What is/are the main feature(s) of interest in your dataset?
#### I am mainly interested in the following features: candidate name and contribution receipt amount, since the most interesting topic is analyzing the amount of contributions received by candidates.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
#### Features like contributor name, contributor city, contributor zipcode, contributor employer, contributor occupation, contribution, election type, candidate's political party can provide further insights of the dataset. We can find relationship between these features and the features I mentioned in the last question. For example, we can try to determine if candidate's political party has an effect on contribution amount.

### Did you create any new variables from existing variables in the dataset?
#### I created a new variable cand_nm_party, which is the candidate's party information(e.g: Democrat, Republican etc) in order to understand the differences among contributions to different parties.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
#### I noticed that the contributions come from many cities but many of them concentrate on one city - Chicago. Similar situation happens to occupation. Additionally, the contributions received by candidates vary significantly, therefore I added a log transformation to make the plot easier to read.


# Bivariate Plots Section
### We can investigate the dataset by looking at two variables. 
#### First, let's find out the amount of contributions received by each candidate.
```{r echo=FALSE, Bivariate_Plots_1}
# Visualize contribution amount for each candidate using boxplot
ggplot(aes(x=cand_nm, y = contb_receipt_amt), 
        data=subset(PC, contb_receipt_amt>0)) +
  geom_boxplot()+           
  scale_y_log10() +
  coord_flip()

# Visualize the total contribution amount for each candidate using bar plot
ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = subset(PC, !is.na(PC$contb_receipt_amt))) +
  stat_summary_bin(aes(y = contb_receipt_amt), 
                   fun.y = 'sum', 
                   geom = 'bar', 
                   data = subset(PC, !is.na(PC$contb_receipt_amt))) +
  coord_flip()
```

#### The boxplots of contribution amount are useful, as they show show 25%, 50% and 75% percentile of the contribution amount for each candidate, along with outliers. Bush had the highest median contribution amount and Clinton had the lowest median contribution amount.

#### Similar to the univariate plot of contribution transactions of each candidate, we are seeing that Clinton received the highest total amount of contribution, followed by Sanders, Cruz and Trump.

#### Another bivariate plot we can create is contributions for each candidate, filled with election type information.
```{r echo=FALSE, Bivariate_Plots_2}
# Visualize the number of contributions received by each candidated, with information of election type
ggplot(aes(x = reorder_size(cand_nm)), data = PC) +
  geom_bar(aes(fill = election_tp)) +
  coord_flip() +
  scale_y_log10()
```

#### Compared to the previous plot for contributions received by candidates, we obtain additional information about how many contributons were made to primary and general election in this plot.

#### We can also investigate amount of contributions made to candidates of different political parties using boxplots.
```{r echo=FALSE, Bivariate_Plots_3}
ggplot(aes(x=cand_nm_party, y = contb_receipt_amt), 
        data=subset(PC, contb_receipt_amt>0)) +
  geom_boxplot()+           
  scale_y_log10()
```

#### The boxplots show that the Libertarian party had the highest median contribution amount, followed by the Green Party, the Republican party and the Democratic party. However, we must keep in mind that the sample sizes for the Libertarian party and the Green party are much smaller.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
#### One obvious pattern is that contribution amounts vary significantly across different candidates in terms of total, median, etc. Another pattern is that in general, more contributions go toward primary election rather than general election. More money is dedicated to make the candiates the party nominee perhaps since only one candidate from each party actually competes in the general election. 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
#### The amount of contributions made to candidates of different political party vary quite a lot, as we see in the boxplots. However, even though the candidates from the Democratic party received the lowest median contribution amount, they received the highest number of contributions, which drives up their total amount of contributions received. It is fair to say that Democratic candidates build their campaigns upon small donations in large quantities.

### What was the strongest relationship you found?
#### Since we only have one numeric variable - contribution amount, we cannot find the correlations.


# Multivariate Plots Section

### We can also investigate the data set by looking at more than 2 variables
#### Based on the previous bar plot of total contribution amount received for each candidate, I created a multivariate plot by filling the data with election type information. 
```{r echo=FALSE, Multivariate_Plots_1}
# Visualize the total contribution amount for each candidate, filled with election type information
ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = PC) +
  stat_summary_bin(aes(y = contb_receipt_amt, fill = election_tp), 
                   fun.y = 'sum', 
                   geom = 'bar') +
  coord_flip()
```

#### We learnt a great amount of information from this plot. In particular, Trump and Clinton are the only candidates with significant contribution amount to their general election campaign. This makes sense considering that they are the two major candidates remained in the general election period. It is also interesting to see that Trump received a lot more contributions during general election than primary election.

#### Another interesting multivariate plot would be the bar plot of total amount of contributions for each candidate, faceted by political party.
```{r echo=FALSE, Multivariate_Plots_2}
# Visualize the total contribution amount for each candidate, faceted by political party
ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = PC) +
  stat_summary_bin(aes(y = contb_receipt_amt), 
                   fun.y = 'sum', 
                   geom = 'bar') +
  facet_wrap(~cand_nm_party) +
  coord_flip()
```

#### This plot is useful because it allows us to compare candidates' total contribution amount to other candidates of their own political party. Within the Democratic party, Clinton received the most contribution amount. And within the Republican party, Trump received the most contribution amount. We can also see that, the Democratic party has fewer candidates but the two main candidates received a lot more contribution amount than candidates from the Republican party.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
#### Seperating the bar plots of total contributions received by candidates into different political party and looking at the boxplots of contribution amounts of candidates from different political parties definitely showed us a clear trend that Democratic candidates take smaller donations in a larger quantities, whereas Republican candidates takes bigger donations but in a smaller quantities.

### Were there any interesting or surprising interactions between features?
#### I thought that the interactions between total contribution amount and election type is quite interesting. It confirms our intuition that only the major candidates who become the party nominees would receive donations for general election.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x=cand_nm_party, y = contb_receipt_amt), 
        data=subset(PC, contb_receipt_amt>0)) +
geom_boxplot()+           
scale_y_log10()+
  ggtitle('Boxplots of Contribution Amount for Candidates from Different Political Parties') +
  xlab('Political Party of Candidate') +
  ylab('Contribution Amount Received (in USD)')
```

### Description One
#### This plot demonstrates the first quantile, median, third quantile and medians contribution amount for candidates of different political party.

### Plot Two
```{r echo=FALSE, Plot_Two}
plot.title = 'Bar Plots of Total Contribution Amount Received for Each Candidate'
plot.subtitle = 'With Election Type Information'

ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = PC) +
  stat_summary_bin(aes(y = contb_receipt_amt, 
                       fill = election_tp), 
                       fun.y = 'sum', 
                       geom = 'bar') +
  coord_flip() +
  ggtitle(bquote(atop(.(plot.title), 
                      atop(italic(.(plot.subtitle)), "")))) +
  ylab('Total Contribution Amount Received (in USD)') +
  xlab('Candidate Name') +
  guides(fill=guide_legend(title="Election Type"))
```

### Description Two
#### This plot demonstrates the total amount of contributions received for each candidate, filled with election type information (general or primary).

### Plot Three
```{r echo=FALSE, Plot_Four}
ggplot(aes(x=cand_nm, y = contb_receipt_amt), 
        data=subset(PC, contb_receipt_amt>0)) +
  geom_boxplot()+           
  scale_y_log10() +
  coord_flip() +
  ggtitle('Boxplots of Contribution Amount Received by Each Candidate') + xlab('Candidate Name') +
  ylab('Contribution Amount Received (in USD)')
```

### Description Three

#### The boxplots show 25%, 50% and 75% percentile of the contribution amount for each candidate, along with outliers.

------

# Reflection
### By doing univariate, bivariate and multivariate analysis, we gained a lot of insights about the political contributions made to 2016 presidential candidates in Illinois, including which candidate received the most contributions and total amount of contributions, summary statistics of contributions for each candidate, differences in contributions across election types and political parties. 

### One major finding is that Democratic candidates received smaller donations but in a larger quantities, whereas the opposite is true for Republican candidates. 

### However, one limitation of this analysis is that since most of the variables are categorical, finding correlations is difficult.

### A way to improve this analysis in the future is to incorporate more demographic information of the contributor, if possible. For example, we can conjecture the gender of the contributor based on the first name, then analyze if the contribution behavior between males and females are different, such as the contribution amount and the candidates they tend to contribute to.